---
title: "Census Dataset Analysis"
preamble: >
  \usepackage[utf8]{inputenc}
  \usepackage[square,numbers]{natbib}
  \usepackage{graphicx}
  \usepackage{enumerate}
  \usepackage{enumitem}
vignette: >
  %\VignetteIndexEntry{COMPAS Analysis - A Causal Approach}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../bib/shai.bib
biblio-style: unsrt
output: >
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
root <- rprojroot::find_root(rprojroot::is_git_root)
r_dir <- file.path(root, "r")
invisible(lapply(list.files(r_dir, full.names = TRUE), source))
invisible(lapply(list.files(file.path(root, "scripts", "helpers"),
                            full.names = TRUE), source))
```

The United States Census of 2018 collected broad information about the US Government employees, including demographic information $Z$ ($Z_1$ for age, $Z_2$ for race, $Z_3$ for nationality), gender $X$ ($x_0$ female, $x_1$ male), marital and family status $M$, education information $L$, and work-related information $R$. 

A data scientist loads the data and performs the following initial analysis:

```{r initial}
data <- get(data("gov_census", package = "faircause"))
data <- as.data.frame(data[seq_len(20000)])
knitr::kable(head(data), caption = "Census dataset.")
tapply(data$salary, data$sex, mean)
```

Therefore, the data scientist observed that male employees on average earn \$14000/year more than female employees, that is
$$ 
E[y \mid x_1] -  E[y \mid x_0] = \$14000.
$$

Following the Fairness Cookbook, the data scientist does the following:

**SFM projection:** the SFM projection of the causal diagram $\mathcal{G}$ of this dataset is given by
$$
\Pi_{\text{SFM}}(\mathcal{G}) = \langle X = \lbrace X \rbrace,  Z = \lbrace Z_1, Z_2, Z_3 \rbrace, W = \lbrace M, L, R\rbrace, Y = \lbrace Y \rbrace\rangle.
$$
She then inputs this SFM projection into the `faircause` R-package,
```{r cookbook}
#| label: fig-census-cookbook
#| fig-cap: Fairness Cookbook on the Census dataset.
#| fig-width: 8
#| fig-height: 6
set.seed(2022)
mdata <- SFM_proj("census")
mdata
fc_census <- fairness_cookbook(data, X = mdata$X, Z = mdata$Z, W = mdata$W,
                               Y = mdata$Y, x0 = mdata$x0, x1 = mdata$x1)

autoplot(fc_census, decompose = "xspec", dataset = "Census")
```

Using these results, she considers the following:

**Disparate treatment:** when considering disparate treatment, she computes $x\text{-DE}^{\text{sym}}_{x}(y \mid x_0)$ and its 95\% confidence interval to be 
$$
x\text{-DE}^{\text{sym}}_{x}(y \mid x_0) = \$9980 \pm \$1049.
$$
The hypothesis $H_0^{(x\text{-DE})}$ is thus rejected, providing evidence of disparate treatment of females.

**Disparate impact:** when considering disparate impact, she notice that Ctf-SE, Ctf-IE and their respective 95\% confidence intervals equal:
$$
	\begin{align}
	    x\text{-IE}^{\text{sym}}_{x}(y \mid x_0) &= \$5126 \pm \$778,\\
	    x\text{-SE}_{x_1, x_0}(y) &= -\$1675 \pm  \$955.
	\end{align}
$$
The data scientist decides that the differences in salary explained by the spurious correlation of gender with age, race, and nationality are not considered discriminatory. Therefore, she tests the hypothesis
	$$H_0^{(x\text{-IE})}: x\text{-IE}^{\text{sym}}_{x}(y \mid x_0) = 0,$$ 
which is rejected, indicating evidence of disparate impact on female employees of the government. The measures computed in this example are visualized in @fig-census-cookbook.
