---
title: "Task 2: Fair Predictions on COMPAS"
output: html_document
bibliography: ../bib/CFA.bib
biblio-style: unsrt
jupyter: python3
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
root <- rprojroot::find_root(rprojroot::is_git_root)
r_dir <- file.path(root, "r")
invisible(lapply(list.files(file.path(root, "scripts", "helpers"),
                            full.names = TRUE), source))
dataset <- "compas"
library(fairadapt)
library(faircause)
library(reticulate)
library(ranger)
library(ggplot2)
library(latex2exp)

RejectOption <- function(prob, cl) {

  rate_cl <- tapply(prob, cl, function(x) mean(x > 0.5))
  fav_cl <- names(rate_cl)[which.max(rate_cl)]

  theta <- 0.5 + seq(0, 0.5, 0.005)
  tv <- vapply(
    theta,
    function(th) {
      mean(prob[cl == fav_cl] > th) - mean(prob[cl != fav_cl] > 1 - th)
    }, numeric(1L)
  )

  theta_min <- theta[which.min(abs(tv))]

  prob[cl == fav_cl] <- prob[cl == fav_cl] > theta_min
  prob[cl != fav_cl] <- prob[cl != fav_cl] > 1 - theta_min

  prob

}


get_mat <- function(src = "compas") {
  
  mat <- switch(src, adult = get_mat_adult(),
                compas = get_mat_compas(),
                census = get_mat_census(),
                berkeley = get_mat_berkeley())
  
  mat
  
}

get_mat_compas <- function() {
  
  
  vars <- c("age", "sex", "juv_fel_count","juv_misd_count",
            "juv_other_count", "priors_count",
            "c_charge_degree", "race", "two_year_recid")
  
  adj.mat <- array(0, dim = c(length(vars), length(vars)))
  colnames(adj.mat) <- vars
  rownames(adj.mat) <- colnames(adj.mat)
  
  # adding the edges to the matrix
  adj.mat[
    c("race", "sex", "age"), c("juv_fel_count", "juv_misd_count",
                               "juv_other_count", "priors_count",
                               "c_charge_degree", "two_year_recid")] <- 1
  adj.mat[c("juv_fel_count", "juv_misd_count", "juv_other_count"),
          c("priors_count", "c_charge_degree", "two_year_recid")] <- 1
  adj.mat["priors_count", c("c_charge_degree", "two_year_recid")] <- 1
  adj.mat["c_charge_degree", "two_year_recid"] <- 1
  
  cfd.mat <- adj.mat
  cfd.mat[, ] <- 0L
  cfd.mat[c("sex", "age"), "race"] <- cfd.mat["race", c("sex", "age")] <- 1L
  
  list(adj.mat, cfd.mat)
  
}

set.seed(2024)
```

In this vignette, we focus on Task 2 (Fair Prediction), and specifically focus on the limitation of the currently found methods in the literature to provide causally meaningful fair predictions. 

## Fair Predictions on COMPAS

A team of data scientists from ProPublica have shown that the COMPAS dataset from Broward County contains a strong racial bias against minorities. They are now interested in producing fair predictions $\widehat{Y}$ on the dataset, to replace the biased predictions. They first load the COMPAS data:
```{r}
# load the data
dat <- get(data("compas", package = "faircause"))
knitr::kable(head(dat), caption = "COMPAS dataset.")

# load the SFM projection
mdat <- SFM_proj("compas")
```
To produce fair predictions, they first experiment with four different classifiers.

#### **(i)** Random Forest without fairness constraints

```{r}
# Method 1: Random Forest
org_dat <- dat
org_dat$two_year_recid <- ranger(two_year_recid ~ ., dat,
                                 classification = TRUE)$predictions

# Method 1: decompose Total Variation
org_tvd <- fairness_cookbook(org_dat, mdat[["X"]], mdat[["Z"]], mdat[["W"]],
                             mdat[["Y"]], mdat[["x0"]], mdat[["x1"]])
org_plt <- autoplot(org_tvd, decompose = "xspec", dataaset = dataset) +
  xlab("") + ylab("")
```

#### **(ii)** Logistic regression trained with _reweighing_ [@kamiran2012data]

```{python, message=FALSE}
#| warning: false
#| message: false
import os

exec(open(os.path.join(r.root, "py", "reweighing_compas.py")).read())
Yhat_rew = reweigh_and_predict(r.dat, r.dat)
```

```{r}
# Method 2: Reweighing by Kamiran & Calders
reweigh_dat <- dat
reweigh_dat$two_year_recid <- as.vector(py$Yhat_rew)

reweigh_tvd <- fairness_cookbook(
  reweigh_dat, mdat[["X"]], mdat[["Z"]], mdat[["W"]], mdat[["Y"]], mdat[["x0"]],
  mdat[["x1"]])

rew_plt <- autoplot(reweigh_tvd, decompose = "xspec", dataset = dataset) +
  xlab("") + ylab("")
```

#### **(iii)** Fair Reductions approach of [@agarwal2018reductions]

```{python}
exec(open(os.path.join(r.root, "py", "reductions_compas.py")).read())
Yhat_red = reduce_and_predict(r.dat, r.dat, 0.01)
```


```{r}
# Method 3: Reductions (Agarwal et. al.)
source_python(file.path(root, "py", paste0("reductions_", dataset, ".py")))

red_dat <- dat
red_dat$two_year_recid <- as.vector(py$Yhat_red)
red_tvd <- fairness_cookbook(
  red_dat, mdat[["X"]], mdat[["W"]], mdat[["Z"]], mdat[["Y"]], mdat[["x0"]],
  mdat[["x1"]]
)

red_plt <- autoplot(red_tvd, decompose = "xspec", dataset = dataset) +
  xlab("") + ylab("")
```

#### **(iv)** Random Forest with _reject-option_ post-processing [@kamiran2012decision]

```{r}
# Method 4: reject-option classification
rjo_prb <- ranger(two_year_recid ~ ., dat, probability = TRUE)$predictions[, 2]
rjo_dat <- dat
rjo_dat$two_year_recid <- RejectOption(rjo_prb, rjo_dat$race)

rjo_tvd <- fairness_cookbook(rjo_dat, mdat[["X"]], mdat[["W"]], mdat[["Z"]],
                             mdat[["Y"]], mdat[["x0"]], mdat[["x1"]])
rjo_plt <- autoplot(rjo_tvd, decompose = "xspec", dataset = dataset) +
  xlab("") + ylab("")
```

### Are the methods successful at eliminating discrimination?

The fair prediction algorithms used above are intended to set the TV measure to $0$. After constructing these predictors, the ProPublica team made use of the Fairness Cookbook, to inspect the causal implications of the methods. Following the steps of the Fairness Cookbook, the team computes the TV measure, together with the appropriate measures of direct, indirect, and spurious discrimination. We can now visualize these decompositions in @fig-compas-fairpred.

```{r}
#| label: fig-compas-fairpred
#| fig-cap: Fair Predictions on the COMPAS dataset.
#| fig-width: 18
#| fig-height: 12
#| echo: false
multi_plot <- function(..., labels = NULL, ncol = 1L) {

  plotlist <- list(...)
  y_range <- lapply(plotlist, function(obj) layer_scales(obj)$y$range$range)
  y_range <- Reduce(rbind, y_range)
  ylims <- c(min(y_range[, 1]), max(y_range[, 2]))

  plotlist <- lapply(plotlist, function(obj) obj + ylim(ylims))

  cowplot::plot_grid(plotlist = plotlist, labels = labels, ncol = ncol,
                     label_size = 20)

}
multi_plot(
  org_plt +
    ggtitle(TeX("Random Forest")),
  rew_plt +
    ggtitle(TeX("Reweighing")),
  red_plt +
    ggtitle(TeX("Reductions")),
  rjo_plt +
    ggtitle(TeX("Reject-Option")),
  labels = c("(i)", "(ii)", "(iii)", "(iv)"), ncol = 2L
)
```

The ProPublica team notes that all methods substantially reduce the $\text{TV}_{x_0,x_1}(\widehat{y})$, however, the measures of direct, indirect, and, spurious effects are not necessarily reduced to $0$, consistent with the Fair Prediction Theorem. 

## How to fix the issue?

To produce causally meaningful fair predictions, we suggest using the `fairadapt` package [@plevcko2020fair; @plevcko2021fairadapt]. In particular, the package offers a way of removing discrimination which is based on the causal diagram. In this application, we are interested in constructing fair predictions that remove both the direct and the indirect effect. Firstly, we obtain the adjacency matrix representing the causal diagram associated with the COMPAS dataset:

```{r}
set.seed(2022)
# load the causal diagram
mats <- get_mat(dataset)
adj.mat <- mats[[1L]]
cfd.mat <- mats[[2L]]

causal_graph <- fairadapt::graphModel(adj.mat, cfd.mat)
layout_matrix <- matrix(c(
  1, 2,  # age
  -1, 2,  # sex
  -2, -2,  # juv_fel_count
  -1, -2, # juv_misd_count
  0, -2, # juv_other_count
  1, -2,  # priors_count
  2, -2,  # c_charge_degree
  -3, 0,  # race
  3, 0    # two_year_recid
), ncol = 2, byrow = TRUE)

plot(causal_graph, layout=layout_matrix, vertex.size = 30)
```

After loading the causal diagram, we perform fair data adaptation:

```{r}
#| label: fig-compas-fairadapt
#| fig-cap: Fair Data Adaptation on the COMPAS dataset.
#| fig-width: 8
#| fig-height: 6
fdp <- fairadapt::fairadapt(two_year_recid ~ ., prot.attr = "race",
                            train.data = dat, adj.mat = adj.mat)

# obtain the adapted data
ad_dat <- fairadapt:::adaptedData(fdp)
ad_dat$race <- dat$race

# obtain predictions based on the adapted data
adapt_oob <- ranger(two_year_recid ~ ., ad_dat,
                    classification = TRUE)$predictions
ad_dat$two_year_recid <- adapt_oob

# decompose the TV for the predictions
dat.fairadapt <- dat
dat.fairadapt$two_year_recid <- adapt_oob

fairadapt_tvd <- fairness_cookbook(
  ad_dat, mdat[["X"]], mdat[["W"]], mdat[["Z"]],
  mdat[["Y"]], mdat[["x0"]], mdat[["x1"]]
)

# visualize the decomposition
fairadapt_plt <- autoplot(fairadapt_tvd, decompose = "xspec", dataset = dataset) +
  xlab("") + ylab("")

fairadapt_plt
```

@fig-compas-fairadapt shows how the `fairadapt` package can be used to provide causally meaningful predictions that remove both direct and indirect effects.
