---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(ggplot2)
```

# [faircause](https://github.com/dplecko/CFA)

<!-- badges: start 
[![Lifecycle](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R build status](https://github.com/dplecko/fairadapt/workflows/build/badge.svg)](https://github.com/dplecko/fairadapt/actions?query=workflow%3Abuild)
[![R check status](https://github.com/dplecko/fairadapt/workflows/check/badge.svg)](https://github.com/dplecko/fairadapt/actions?query=workflow%3Acheck)
[![pkgdown build status](https://github.com/dplecko/fairadapt/workflows/pkgdown/badge.svg)](https://github.com/dplecko/fairadapt/actions?query=workflow%3Apkgdown)
[![covr status](https://github.com/dplecko/fairadapt/workflows/coverage/badge.svg)](https://github.com/dplecko/fairadapt/actions?query=workflow%3Acoverage)
[![Codecov test coverage](https://codecov.io/gh/dplecko/fairadapt/branch/main/graph/badge.svg?token=8A0EL5N4RE)](https://app.codecov.io/gh/dplecko/fairadapt)
<!-- badges: end -->

The R-package `faircause` can be used for performing Causal Fairness Analysis and implements the methods described in the paper [Causal Fairness Analysis (Plecko & Bareinboim, 2024)](https://causalai.net/r90.pdf). We refer you to the manuscript for full theoretical details. In this repository, you will find a range of examples that demonstrate how to use Causal Fairness Analysis in practice.

### Suggested Citation

To cite the paper, please use the following:
```
@article{plecko2024CFA,
  title={Causal Fairness Analysis: A Causal Toolkit for Fair Machine Learning},
  author={Ple{\v{c}}ko, Drago and Bareinboim, Elias},
  journal={Foundations and Trends{\textregistered} in Machine Learning},
  volume={17},
  number={3},
  pages={304--589},
  year={2024},
  publisher={Now Publishers, Inc.}
}
```

### Installation

You can install `faircause` from this Github repository by using the [devtools](https://cran.r-project.org/web/packages/devtools/index.html) package:

```{r, eval = FALSE}
devtools::install_github("dplecko/CFA")
```

Please note that `faircause` is still under development (currently in version `0.2.0`) and any debug reports or suggested fixes are welcome.

### How to use CFA

A number of vignettes demonstrating how to use the package can be found on our [Github pages](https://dplecko.github.io/CFA/).

### Want to learn more about Causal Fairness Analysis?
For those interested in learning more about CFA, we suggest the following resources:

<ol style="list-style-type: lower-roman; counter-reset: list;">
  <li> Reading the Causal Fairness Analysis paper, found <a href="https://causalai.net/r90.pdf">here</a>,</li>
  <li> Follow the series of <a href="https://www.cs.columbia.edu/~dplecko/#teaching">lectures on CFA</a> which were part of the COMSW-4775 course at Columbia Computer Science,</li>
  <li> Check our <a href="https://fairness.causalai.net/">ICML 2022 Tutorial</a>.</li>
  <li> Check the vignettes on <a href="https://dplecko.github.io/CFA/">Github pages</a> that demonstrate how to perform Causal Fairness Analysis in practice.</li>
</ol>
